Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-09-10 doc/response_to_reviewer_R1.Rmd

**RESPONSE: Thanks for this comment. We have now included the long-term mean trend along with the long-term standard deviation to for Fig 1a to show the deviation more clearly. Fig 1b now follows more closely, showing the deviation in SD units from the long-term mean.**

![](display/Fig1.png)

•	It confuses many acronyms. It's necessary to rewrite their description at least once in the results section, e.g. CII doesn't appear in this section.

**RESPONSE: We have now carefully reviewed the text and rewritten the description of acronyms (TWI and CII) in the results section to make it clearer for readers.**

@@

•	Predictions (i–iii) are clearly formulated, but the Results section is at times overly dense with statistical detail. It would be helpful to present more explicitly which findings support each hypothesis.

**RESPONSE: Thanks for this suggestion. We have now revised the Results section to remove some redundant statistical details that are reported in figures. We have also reorganised some text so that the statements that correspond to each prediction are highlighted at the start of each section. We hope that the results and their alignment with the predictions are clearer now.**

@@

•	Since the study is based on only one site, it is relevant to improve the discussion section with other already published study sites (add more references), mainly link the global implications, such mortality rate and vulnerability in tropical forests. See https://urldefense.proofpoint.com/v2/url?u=https-3A__doi.org_10.1126_science.adq6607&d=DwIFaQ&c=009klHSCxuh5AI1vNQzSO0KGjl4nbi2Q0M1QLJX9BeE&r=6fH468T58at0TtM59sMf109A_dRgo6XTfGMOVX5Ewrw&m=fPE_4y1UPsTl9St_Y44IUO7VIaQJOJIUiHtK15-Nr7g4UiVKbtYQ-TE0043SLAj6&s=SxqVZgFkbqAZ0-VEvdQntXWpGEUu9lr25rgxQtwaVdw&e=

*RESPONSE:*

•	There was something missing to discuss about the species in particular, some that have been more affected than others and why, especially those in conservation categories.

**RESPONSE: Thanks for this suggestion. We have now included a supplementary table (Table S1) that details the response of specific species to specific drought events. This summary highlights our main message that the species responses are dissimilar across drought events. This table also includes available information on the IUCN categories of the species analysed (accessed in September 2025) where information was avaialable. Two of the species analysed are "Endangered", including *Afzelia xylocarpa*, the species that was discussed. There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities. This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions.**

@@

RESPONSE:

Minor comments:

55: Could you briefly explain the three climatically distinct droughts?

**RESPONSE: The three climatically distinct droughts are summarised in Table 1. In summary, 2010 was a moderate drought that peaked in the dry season, 2015 was a severe drought that peaked in the wet season and 2020 was a moderate drought that peaked in the transition from dry to wet season. After moving the drought descriptions to the methods section, we have added this summary to set context for the results section.**

@@

**RESPONSE: Thanks for this comment. We have revised Figure 2 to remove species lines. Originally, we had chosen to include the top 10 species in the plot to illustrate both synchrony and divergence in species responses, but the results text reported on growth and sensitivity values across all species considered (n=30), shown in the Supplementary Appendix. We agree that this misalignment in species choice was confusing and have removed it from Figure 2 and referenced the relevant figure in the results text.**

![](display/Fig2_alt2.png)

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

**RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity. In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees. We have now clarified this in the text which now reads:**

"Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity."

### 2025-09-10 doc/manuscript_hkk_drought_sensitivity.Rmd
updated reviewer comments and manuscript

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 5036  
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. -->
<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

@@

Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.

Species median sensitivities to the three droughts were not correlated with each other, supporting prediction *i*. 
However on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).

@@

<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction *ii*.
Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

## Microenvironment and individual effects {-}

Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).

TWI had different effects across droughts, supporting prediction *iii*. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`). 
The interaction between TWI and deciduousness was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`) but was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) or 2020, (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2020],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2020],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2020],2)`)
<!-- In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.  -->
Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b). 
In 2020, predicted sensitivities of deciduous trees were more positive than evergreen trees and this effect was observable in drier sites (Fig `r fig_species`b).

@@

Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction *iii*.
Crown exposure was associated with greater growth declines in 2010.
In the same models describd above, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`). 
Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 and 2020 (Fig `r fig_dag`b).

<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure across all years; median effect and 90% CI for 2010, 2015 and 2020 respectively were `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`(`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a). 
Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, and not significant in 2015 or 2020 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.

>>>>

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 4999   
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

@@

<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
In summary, the three climatically distinct droughts can be characterised thus: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Table 1).

Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. -->
<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

@@

Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction *i* (Fig S5, Fig S6). 
Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*. 
Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S5). Median species sensitivities to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->

@@

<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction *ii*.
Deciduous species were more drought sensitive than evergreen species in 2015 <!--(Fig `r fig_species`a, Pearson's correlation coefficient r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`)-->(Fig `r fig_species`a).
There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Models of all trees with Topographic Wetness Index (TWI) and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

## Microenvironment <!--and individual -->effects {-}

<!-- Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).  -->
TWI had different effects across droughts, supporting prediction *iii*. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`). 
The interaction between TWI and deciduousness was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`) but was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) or 2020, (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2020],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2020],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2020],2)`)
<!-- In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.  -->
Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b). 
In 2020, predicted sensitivities of deciduous trees were more positive than evergreen trees and this effect was observable in drier sites (Fig `r fig_species`b).

@@

Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction *iii*.
Crown Illumination index (CII) was associated with greater growth declines in 2010.
In the same models described above, the simplex parameter for the ordered factor CII in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`). 
Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity (Fig `r fig_dag`b).

<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher CII across all years; median effect and 90% CI for 2010, 2015 and 2020 respectively were `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`(`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a). 
Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, and not significant in 2015 or 2020 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models. Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020. Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size. d) Modelled sensitivity across CII classes in 2010, 2015 and 2020.
